\propertitle

### 2019-07-20 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-13T11:35:37+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\mynote{Move this to intro?}We want to stress the usefulness of making
quantified guesses about the full-network activity. First of all, this
seems to be the primary idea behind recording a sample. Second, it allows
us to make comparisons across experimental sessions; such comparisons would
me difficult or meaningless if made with the recorded samples, which
generally comprise non-overlapping sets of neurons and differ in size.

@@

The distribution for the full-network is sharper than the measured
frequency distribution for the sample; the sharper the larger $N$ is. Most
remarkably, it has two distinct low-activity modes for the first data set.
But also the second data set presents, upon closer inspection, a small
shoulder on the right of the mode, suggestive of two activity regimes. We
frame no hypotheses about the biological cause of these two modes (they
could stem from the presence of different kinds of cells or modules). These
features are clearly not present in the sample or in the maximum-entropy
distribution at the sample level. The application of the probability
calculus thus reveals interesting possible features of the full network.

@@

\section{Assumptions and corrections:\\beyond the maximum-entropy approximation}
\label{sec:assumptions_formulae}

In \sect~\ref{sec:model} we mentioned that maximum-entropy distributions
come from a zeroth-order Laplace approximation of the densities obtained
from the probability calculus. Zeroth-order means that we are simply
considering the mode of such densities. Let's be more concrete.

Consider the question: What is the relative-frequency distribution for the
different activity levels for the full-network, across the recorded bins?
Call this distribution $(\yFF_{\yA}) \defs \yF$: activity level $\yA$
appeared in $T\times\yFF(\yA)$ out of $T$ time bins. We don't know $\yF$,
and our beliefs about its possible values have a distribution
$\pf(\yF\|\yD\yI)$ conditional on data $\yD$ and prior knowledge $\yI$,
which we approximate with a continuous distribution
\begin{equation}
  \label{eq:density_F}
  \pf(\yF \| \yD \yI)\;\di\yF.
\end{equation}
In appendix*** we give a summary of how this distribution is derived. Note that
this is a probability distribution over frequency distributions, so we must
be careful with terminology to avoid confusion.

Now consider the related question: What is the activity $\yA_{t}$ of the
full network at some specific time bin $t$? If we don't know the exact
sequence of activities during the recording then we don't know $\yA_{t}$;
if we know the frequency distribution $\yF$ of the activity levels, then
our \dob\ about $\yA_{t}$ is $\yFF_{\yA_{t}}$, out of combinatorial
reasons. But if we are also uncertain about $\yF$, with
belief distribution~\eqref{eq:density_F}, then our \dob\ about $\yA_{t}$ by
marginalization is
\begin{equation}
  \label{eq:probability_AT}
  \pf(\yA_{t} \| \yD \yI) =
  \int \yFF_{\yA_{t}}\;   \pf(\yF \| \yD \yI)\;\di\yF;
\end{equation}
our belief distribution for $\yA_{t}$ is thus simply the mean
$\int \yF\, \pf(\yF \| \yD \yI)\,\di\yF$.

Formulae~\eqref{eq:density_F} and \eqref{eq:probability_AT} show that the
most probable frequency distribution $\yF$ (if it's unique) and the
probability distribution for $\yA_{t}$ are generally different: the former
is the mode of $\pf(\yF \| \yD \yI)$, the latter its mean.

If the probability density $\pf(\yF \| \yD \yI)$ is extremely peaked at some
frequency distribution $\yFm$, it can be approximately treated as a delta
density,
\begin{equation}
  \label{eq:approx_delta_pF}
  \pf(\yF \| \yD \yI)\;\di\yF \approx \delt(\yF-\yFm)\;\di\yF,
\end{equation}
and consequently the probability for the activity at bin $t$, by
\eqn~\eqref{eq:probability_AT}, is simply the mode $\yFm$ itself:
\begin{equation}
  \label{eq:approx_delta_pA}
  \pf(\yA_{t} \| \yD \yI) \approx \yFFm_{\yA_{t}}.
\end{equation}
The maximum-entropy distribution is exactly this mode $\yFm$, for
\emph{particular} kinds of probability densities $\pf(\yF \| \yD \yI)$
which we'll discuss shortly. Formulae~\eqref{eq:approx_delta_pF}
and~\eqref{eq:approx_delta_pA} show why the maximum-entropy distribution is
an approximate answer to two different questions.

If the probability density $\pf(\yF \| \yD \yI)$ is not sufficiently
peaked, however, its mode and mean can be quite different. The
maximum-entropy approximation can then be inadequate for two reasons.
First, despite being the mode it can be a poor representative of all the
possible frequency distributions around the mode. Second, it can be a bad
approximation of the probability distribution~\eqref{eq:probability_AT}.

Let's therefore examine the full density $\pf(\yF \| \yD \yI)$ given by the
probability calculus. Its derivation is given in appendix***; here we
 discuss its mathematical features and examine the answers it
provides. Its expression,
\begin{equation}
  \label{eq:density_proportional}
  \pf(\yF \| \yD \yI) \propto \exp\biggl[
-  \frac{\abs{\ycm\yF - \yc}^{2}}{2\ycs^{2}}
-  \frac{\abs{\yd\yF}^{2}}{2\yds^{2}}
-\yL\;\ysh(\yF;\yR)\biggr],
\end{equation}
consists of three terms:

\textbullet\ The density
\begin{equation}
\exp[-\yL\;\ysh(\yF;\yR)]\label{eq:entropic_prior}
\end{equation}
expresses our pre-data
beliefs about the possible frequency distributions. Here
$\ysh(\yF;\yR)\defd \sum_{\yA}\yFF_{\yA}\ln(\yFF_{\yA}/\yRR_{\yA})$ is the
relative entropy or discrimination information
\citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}, $\yR$ is a
reference distribution, and $\yL$ a positive parameter. This density
expresses a belief proportional to the number of ways in which the
frequency distribution $\yF$ can be realized if we assume that the
activities appear with frequencies $\yR$ in the long run, as can be seen
using Stirling's approximation. For $\yR$ we choose the uniform
distribution, but any biologically sensible distribution, such as one
decreasing exponentially with the activity, leads to the same final
results. The parameter $\yL$ roughly represents the number of data points
necessary to change our initial belief, and we set it equal to $50$. This
density is similar to an \enquote{entropic prior}
\citep{rodriguez1991,skilling1998,catichaetal2004,portamana2017}[see
esp.][\sect~3]{neumann2007}.

\textbullet\ The previous density is corrected by the normal density
$\exp\Bigl(-\frac{\abs{\yd\yF}^{2}}{2\yds^{2}}\Bigr)$, where the matrix
$\yd$ is the discrete equivalent of a fourth-order derivative
operator. %ref?
This density expresses our belief, hinging on biological arguments, that
the frequency distribution should be somewhat smooth, without huge
discontinuities between one activity level and the next. The fourth
derivative is chosen because it still allows for the presence of modulated
maxima and minima more than derivatives of lower order do. The standard
deviation $\yds$ is such that the smoothness of the maximum-entropy
distribution, taken as a touchstone, is within two standard deviations.

\textbullet\ The normal density
$\exp\Bigl(-\frac{\abs{\ycm\yF - \yc}^{2}}{2\ycs^{2}}\Bigr)$ is the
likelihood of the frequency $\yF$ in view of our data. The matrix $\ycm$
expresses the linear constraints on $\yF$, and the tuple $\yc$ the measured
values of such constraints. For example, if we consider the full frequency
distribution $\yf$ of the sample, then $\ycm$ effects the marginalization
to a sample of size $n$ through a hypergeometric distribution (sampling
without replacement):
\begin{equation}
  \label{eq:constraint_hypergeometric}
\ycmm_{\ya,\yA} \defd
  \binom{n}{n\ya}\binom{\yNv-n}{\yNv \yA-n\ya}\binom{\yNv}{\yNv \yA}^{-1}
\end{equation}
and $\yc \defd \yf$ are the measured frequencies. If we only consider some
factorial moments, instead, then
\begin{equation}
  \label{eq:constraint_hypergeometric}
  \ycmm_{m,\yA} \defd
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}
\end{equation}
as for \eqn~\eqref{eq:equal_m_factorialmoments}, and $\yc$ is the tuple of
measured factorial moments. The standard deviations $\ycs$ reflect our
beliefs about the measurement results, which can come for example from
the spike-sorting procedure. Their values are discussed in appendix***.

The post-data belief density~\eqref{eq:density_proportional} has a
maximum-entropy distribution as a mode approximation when $\ycs$ is small,
$\yL$ is large, and $\ycs L$ is small: the mode will be the constrained
$\yF$ that minimizes the relative entropy $\ysh(\yF;\yR)$. It should be
noted that a pre-data belief different from the entropic
one~\eqref{eq:entropic_prior} will lead to a different mode approximation.
For example, a Dirichlet density (which is equivalent
to~\eqref{eq:entropic_prior} with $\yF$ and $\yR$ exchanged) would lead to
a maximum-entropy distribution with \emph{Burg's} entropy \citep{burg1975}
instead of Shannon's \citep{jaynes1986d_r1996,portamana2009}.

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/MC_plot1000.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/MC_marginal1000.pdf}%
\caption{Entropic prior}
\label{fig:MC1000}
\end{figure}
% mctests3.nb
% mcfd_6mom_unif_afssd40st0np_cok_tr_N1000_t50_L50_T0.333333333333333_s60000_a15758.0500981316
%
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/MC_plot_unif1000.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/MC_marginal_unif1000.pdf}%
\caption{Prior uniform in $\di\yF$}
\label{fig:MC1000}
\end{figure}
% mctests3.nb
% mcfd_6mom_uunif_afssd40st0np_cok_tr_N1000_t50_L0_T0.333333333333333_s60000_a15758.0500981316
%
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/MC_plot_dirich1000.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/MC_marginal_dirich1000.pdf}%
\caption{Dirichlet prior (entropic with reversed relative entropy)}
\label{fig:MC1000}
\end{figure}
% mctests3.nb
% mcfd_6mom_iunif_afssd40st0np_cok_tr_N1000_t50_L1_T0.333333333333333_s60000_a15758.0500981316
%
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/MC_plot5000.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/MC_marginal5000.pdf}%
\caption{Entropic prior}
\label{fig:MC5000}
\end{figure}
% mctests3.nb
% mcfd_6mom_unif_afssd32s_cok_tr_N5000_t180_L50_T0.333333333333333_s25800_a329212306.071189
The properties of the belief density~\eqref{eq:density_proportional} in the
case of our first data set are shown in \fig~\ref{fig:MC1000} for
$N=1\,000$ and in \fig~\ref{fig:MC5000} for $N=5\,000$. Only the first six
factorial moments were considered as our data, but it turns out that they
are informationally sufficient, leading to practically the same result as
specifying the full frequency distribution for the sample.

The top plots in the figures show the mean
$\int \yF\, \pf(\yF \| \yD \yI)\,\di\yF$
(\textcolor{mypurpleblue}{\textbf{------}}), which is also our belief about the
activity at any single time bin, \eqn~\eqref{eq:probability_AT}, is clearly
different from the maximum-entropy mode. The latter is therefore not a good
approximation for the full density~\eqref{eq:density_proportional}, which
is quite broad, as can be seen from the ten samples
(\textcolor{mygreen}{------}) drawn from it. This broadness is not
surprising: we're making guesses about thousands of neurons from the
observation of fewer than 100. Yet, the mean and the maximum-entropy mode
are qualitatively similar. In fact, for $N=1\,000$ the second peak of the
mean is even higher than the corresponding maximum-entropy one. The samples
also show that generally two high-frequency regions of activity are
expected, especially for $N=5\,000$.

The bottom plots in the figures show our belief distributions for the
marginal frequency of each activity level, $\pf(\yFF_{\yA} \| \yD \yI)$,
with the means (\textcolor{mypurpleblue}{\textbf{------}}), medians
(\textcolor{mygreen}{\textbf{--~--~--}}), and 16\% and 84\% quantiles
(\textcolor{myyellow}{$\bm{\cdot \cdot \cdot \cdot}$} corresponding to one
standard deviation for a normal distribution). The coincidence of means
and medians indicate that these marginals are very close to normal
distributions. This happens thanks to the marginalization, which integrates
out $N-2$ quantities. Note, however, that the full distribution for $\yF$
is far from normal, as indicated by the pronounced differences between mode
and mean.

@@

We hurry to say that the description just given is just \emph{one} way to
quantify and characterize cooperativity and functional grouping, not
\emph{the only} way. It can surely be criticized from many points of view.
Yet, it is quantitative and bears a more precise meaning than an undefined,
vague notion of \enquote{cooperativity}. Two persons who apply this
procedure to the same data will obtain the same numbers. Different
protocols can be based on the \me\ method, for instance protocols that take
into account the activities or pair activities of specific neurons rather
than network averages, or even protocols that take into account time
dependence.
\textcolor{white}{If you find this you can claim a postcard from me.}

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-20T12:48:31+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\mynote{Move this to intro?}We want to stress the usefulness of making a
quantified guess about the full-network activity. First of all, this guess
about a brain area seems to be the primary idea behind recording a sample
from that area. Second, it allows us to make comparisons across
experimental sessions; such comparisons would me difficult or meaningless
if made with the recorded samples, which generally comprise non-overlapping
sets of neurons and differ in size. Third, only at the full-network level
can we meaningfully assess whether correlations of some order are
informationally sufficient: as explained in
\sect~\ref{sec:marginalization}, any such assessment at the sample level
leads to conclusions opposite to the ones we think we're reaching.

@@

The distribution for the full-network is sharper than the measured
frequency distribution for the sample; the sharper the larger $N$ is. Most
remarkably, it has two distinct low-activity modes for the first data set.
But also the second data set presents, upon closer inspection, a small
shoulder on the right of the mode, suggestive of two activity regimes. We
frame no hypotheses about the biological cause of these two modes (they
could stem from the presence of different kinds of cells or modules). These
features are clearly not present in the sample or in the maximum-entropy
distribution at the sample level. The application of the probability
calculus thus reveals interesting possible features of the full network.
\textcolor{white}{If you find this you can claim a postcard from us.}

@@

@@

We hurry to say that the description just given is just \emph{one} way to
quantify and characterize cooperativity and functional grouping, not
\emph{the only} way. It can surely be criticized from many points of view.
Yet, it is quantitative and bears a more precise meaning than an undefined,
vague notion of \enquote{cooperativity}. Two persons who apply this
procedure to the same data will obtain the same numbers. Different
protocols can be based on the \me\ method, for instance protocols that take
into account the activities or pair activities of specific neurons rather
than network averages, or even protocols that take into account time
dependence.
